The role of cognateness in native spoken word recognition

### 2025-07-28 docs/appendix.qmd
Mask preprint

```{r fig-appendix-diagnostics}
#| label: fig-appendix-diagnostics
#| fig-width: 9
#| fig-height: 9
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Cognateness",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Cognateness × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

@@

```{r fig-coefs}
#| label: fig-coefs
#| fig-width: 9
#| fig-height: 5
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Similarity",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Similarity × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

>>>>

```{r}
#| label: fig-appendix-diagnostics
#| fig-cap: "Traceplot of MCMC posterior samples of the fixed regression coefficients of the models in Experiments 1, 2, and 3."
#| fig-width: 9
#| fig-height: 9
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Cognateness",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Cognateness × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

@@

```{r}
#| label: fig-coefs
#| fig-cap: "Posterior distribution of fixed regression coefficients of the models in Experiments 1, 2, and 3."
#| fig-width: 9
#| fig-height: 5
var_labels <- c(
  "b_Intercept" = "Intercept",
  "b_freq_zipf_2_std" = "Frequency",
  "b_lv_std" = "Similarity",
  "b_neigh_n_h_std" = "CLPN",
  "b_neigh_n_h_std:lv_std" = "Similarity × CLPN",
  "b_group1" = "cat-ENG vs. spa-ENG"
)

### 2025-07-28 docs/manuscript.qmd
Mask preprint

---
abstract:  "The present study characterises homophonic translation---a phenomenon by which non-native speech activates words or phrases in a listener's native language---from a language processing perspective. Across three studies, English and Spanish-native monolingual adults completed a translation elicitation task in which participants listened to words in an unfamiliar language (Catalan or Spanish), and were asked to type their best-guess translation in their native language. Our findings suggest that that listening to speech in an unfamiliar language triggers equivalent dynamics of lexical selection than native speech. In particular, phonological similarity between the presented words and their correct translations interacted with phonological neighbourhood density: participants correctly translated words with high phonological similarity with their correct translation only when few other words in the native language shared phonological similarity with the presented word."
---
```{r}
#| label: setup
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(readxl)
library(ggplot2)
library(ggExtra)
library(patchwork)
library(tinytable)
library(brms)
library(tidybayes)
library(stringdist)

@@

One exception is the study by @otake2007interlingual, who analysed 194 instances of Soramimi from the Japanese TV show *Soramimi hour*, broadcasted between 1992 and 2007, in which examples of Soramimi were presented to the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. While the degree to which phonetic features of the presented English lyrics were in the translation varied considerably, the author identified tree main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /kɹaɪ̯/ to *kurai* /kɯ̟ɾa̠i/ [dark]), deletion (e.g., *go* /ɡoʊ/ to *go* /go/ [*Go*, a board game], and alternation (e.g., *low* /loʊ/ to *rou* /ɾo̞ː/ [wax]). These findings suggests that homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to Japanese phonology and phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

Though rarely, homophonic translations sometimes preserve the meaning of the original word or phrase, even if the listener is completely unfamiliar with the language being spoken3. For example, an English native listening to Dutch for the first time might encounter the Dutch word /ˈpɪŋ.(ɡ)ʋɪn/ (*pinguin*) and be able to activate its correct English translation /ˈpɛŋɡwɪn/ (penguin), given how phonologically similar both word-forms are. This phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan than English words by capitalising solely on phonological similarity.

@@

All materials, data, and code used in this study are hosted in the Open Science Framework [https://osf.io/9fjxm/?view_only=751d6ab27ffd4dd4856e4c07c4372991](https://osf.io/9fjxm/?view_only=751d6ab27ffd4dd4856e4c07c4372991) and a GitHub repository [https://github.com/gongcastro/translation-elicitation.git](https://github.com/gongcastro/translation-elicitation.git), along with additional notes. For reproducibility, a Docker image of the RStudio session is available on DockerHub ([LINK REDACTED FOR ANONIMITY]).

@@

We collected data from `r n_participants` British English-native adults living in the United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (£5 compensation) and SONA (compensation in academic credits). Participants gave informed consent before providing any data. The study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.

```{r}
#| label: tbl-participants
#| tbl-cap: "Participant details. *N* indicates the number of included participants, with excluded participants in parenthesis."
tbl_data <- participants |>
  mutate(
    exp = case_when(
      source == "Experiment" & group != "cat-SPA" ~ "Experiment 1",
      source == "Experiment" & group == "cat-SPA" ~ "Experiment 2",
      source == "Questionnaire" ~ "Experiment 3",
    )
  ) |>
  summarise(
    n = n(),
    n_excluded = sum(!valid_status == "Valid"),
    across(age, lst(mean, sd, min, max)),
    l2_lst = list(l2),
    .by = c(group, exp)
  ) |>
  mutate(
    l2_lst = map(
      l2_lst,
      function(x) {
        y <- table(x)
        y <- y[names(y) != "None"]
        y <- paste0(names(y), " (", y, ")")
        return(y)
      }
    ),
    n_excluded = paste0(n, " (", n_excluded, ")"),
    age_mean_sd = paste0(round(age_mean, 2), " (", round(age_sd, 2), ")"),
    age_min_max = paste0(age_min, "--", age_max)
  ) |>
  select(group, n_excluded, age_mean_sd, age_min_max, l2_lst)

@@

In each trial, participants listened to one audio file, which contained a single word. Audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, range = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Spanish audio files was `r round(durations$spa_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, range = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

@@

```{r}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli details."
tbl_data <- stimuli |>
  drop_na(freq_zipf_2, lv, neigh_n_h) |>
  summarise(
    across(c(freq_zipf_2, lv, neigh_n_h), list(mean, sd, min, max)),
    .by = c(group)
  )

@@

tbl_data |>
  tt(
    width = c(
      0.16,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07
    )
  ) |>
  format_tt(
    j = 2:ncol(tbl_data),
    digits = 2,
    num_zero = TRUE,
    num_fmt = "decimal"
  ) |>
  group_tt(
    j = list(
      "Lexical frequency" = 2:5,
      "Phonological simlarity" = 6:9,
      "CLPN" = 10:13
    ),
  )
```

@@

$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{Similarity}_i +  \beta_{3[p]} \text{CLPN}_i + \beta_{4[p]} \text{Group}_i + \\
&\beta_{5[p]} (\text{Similarity}_i \times \text{CLPN}_i) + \beta_{6[p]} (\text{Similarity}_i \times \text{Group}_i) + \\
& \beta_{7[p]} (\text{CLPN}_i \times \text{Group}_i) + \beta_{8[p]} (\text{Similarity}_i \times \text{CLPN}_i \times \text{Group}_i) \\
\beta_{0-8[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-8[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{HalfCauchy}(0, 0.1) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(8) \\
\end{aligned}
$$ {#eq-1}

@@

```{r}
#| label: fig-epreds-1
#| cache: false
#| fig-height: 5
#| fig-width: 11
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
get_epreds <- function(
  model,
  data,
  n = 100,
  lv = seq(0, 1, length.out = 100),
  neigh_n_h = c(0, 2, 4, 8, 12),
  ...
) {
  lv_std <- (lv - mean(data$lv, na.rm = TRUE)) / sd(data$lv, na.rm = TRUE)
  neigh_n_h_std <- (neigh_n_h - mean(data$neigh_n_h, na.rm = TRUE)) /
    sd(data$neigh_n_h, na.rm = TRUE)
  knowledge <- unique(model$data$knowledge)
  confidence <- unique(model$data$confidence)
  freq_zipf_2_std <- 0
  group <- unique(model$data$group)
  experiment <- unique(model$data$experiment)
  nd <- expand_grid(
    freq_zipf_2_std,
    neigh_n_h_std,
    lv_std,
    knowledge,
    confidence,
    group,
    experiment
  )
  epreds <- tidybayes::add_epred_draws(nd, model, re_formula = NA, ...)
  return(epreds)
}

@@

We collected data from `r n_participants` Spanish native adults living in Spain (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants in Spain were contacted via announcements at the University campus(es), and were compensated €5 or an Amazon voucher for the same value. Participants gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar (2020/9080/I).

@@

```{r}
#| label: fig-epreds-2
#| cache: false
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
#| fig-height: 3.5
#| fig-width: 10
lv_std <- (lv - mean(dataset_2$lv_std, na.rm = TRUE)) /
  sd(dataset_2$lv_std, na.rm = TRUE)
neigh_n_h <- c(0, 2, 4, 8, 12)
epreds_2 <- get_epreds(exp_2_m0, dataset_2, neigh_n_h = neigh_n_h, lv = lv)
epreds_2$neigh_n_h_std <- factor(
  epreds_2$neigh_n_h_std,
  levels = unique(epreds_2$neigh_n_h_std),
  labels = paste0(neigh_n_h, " neighbours"),
  ordered = TRUE
)

@@

We collected data from `r n_participants` British English native adults living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. Stimuli were the same lists of Catalan and Spanish stimuli as in Experiment 1.

@@

Future studies may consider comparing the suitability of other measures of phonological similarity that take into account finer-grained phonetic or prosodic cues present in the acoustic signal presented to participants. Phonological transcriptions like X-SAMPA are abstract representations that ignore non-phonological contrasts (e.g., phones), and consider different symbols as completely different phonemes, disregarding the fact that some phonemes share more phonetic features than others. Additionally, prosodic cues such as lexical stress, which our measure of phonological similarity does not include, might also provide participants further information about the correct translation of the presented word-forms. Altogether, it is likely that participants in the present study were able to exploit additional cues in the acoustic signal to provide correct translations. Future studies may explore the relative contribution of such cues in the occurrence of homophonic translation. Another possible limitation of the present study is the limited size and diversity of the all of which had high lexical frequency and low age-of-acquisition. To better characterise the factors that guide listeners’ ability to match words from an unfamiliar language with words in their native language lexicon, words with more varying levels of difficulty should be included in future studies.

>>>>

---
abstract:  "Homophonic translation is a popular perceptual phenomenon by which non-native speech activates words or phrases in a listener's native language. In this study, we characterise homophonic translation from a language processing perspective. Across three studies, English and Spanish-native monolingual adults completed a translation elicitation task in which participants listened to words in an unfamiliar language (Catalan or Spanish), and were asked to type their best-guess translation in their native language. Our findings suggest that that listening to speech in an unfamiliar language triggers equivalent dynamics of lexical selection than native speech. In particular, phonological similarity between the presented words and their correct translations interacted with phonological neighbourhood density: participants correctly translated words with high phonological similarity with their correct translation only when few other words in the native language shared phonological similarity with the presented word."
---
```{r}
#| label: setup
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(purrr)
library(readxl)
library(ggplot2)
library(ggExtra)
library(patchwork)
library(tinytable)
library(brms)
library(tidybayes)
library(stringdist)

@@

One exception is the study by @otake2007interlingual, who analysed 194 instances of Soramimi from the Japanese TV show *Soramimi hour*, broadcasted between 1992 and 2007, in which examples of Soramimi were presented to the audience for comedy purposes. These instances consisted of homophonic translations of English song lyrics to words and phrases in Japanese. While the degree to which phonetic features of the presented English lyrics were preserved in the translation varied considerably, Otake identified tree main phonological processes that could explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /kɹaɪ̯/ to *kurai* /kɯ̟ɾa̠i/ [dark]), deletion (e.g., *go* /ɡoʊ/ to *go* /go/ [*Go*, a board game], and alternation (e.g., *low* /loʊ/ to *rou* /ɾo̞ː/ [wax]). These findings suggests that homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to Japanese phonology and phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic].

Though rarely, homophonic translations sometimes preserve the meaning of the original word or phrase, even if the listener is completely unfamiliar with the language being spoken. For example, an English native listening to Dutch for the first time might encounter the Dutch word /ˈpɪŋ.(ɡ)ʋɪn/ (*pinguin*) and be able to activate its correct English translation /ˈpɛŋɡwɪn/ (penguin), given how phonologically similar both word-forms are. This phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan words than English words via homophonic translation, that is, by capitalising solely on phonological similarity.

@@

All materials, data, and code used in this study are hosted in the Open Science Framework ([https://osf.io/9fjxm/?view_only=751d6ab27ffd4dd4856e4c07c4372991](https://osf.io/9fjxm/?view_only=751d6ab27ffd4dd4856e4c07c4372991)) and a GitHub repository ([https://anonymous.4open.science/r/translation-elicitation-8BCE](https://anonymous.4open.science/r/translation-elicitation-8BCE) [FOR ANONYMITY, LINK TEMPORARILY REDIRECTED TO AN ANONYMOUS MIRROR OF THE GITHUB REPOSITORY]), along with additional notes. For reproducibility, a Docker image of the RStudio session is available on DockerHub ([LINK REDACTED FOR ANONIMITY]).

@@

We collected data from `r n_participants` British English-native adults living in the United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (£5 compensation) and SONA (compensation in academic credits). Participants gave informed consent before providing any data. The study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the [ETHICAL COMMITTEE REDACTED FOR ANONYMITY]. Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection.

```{r}
#| label: tbl-participants
#| tbl-cap: "Participant information. *N* indicates the number of included participants, with excluded participants between parentheses."
tbl_data <- participants |>
  mutate(
    exp = case_when(
      source == "Experiment" & group != "cat-SPA" ~ "Experiment 1",
      source == "Experiment" & group == "cat-SPA" ~ "Experiment 2",
      source == "Questionnaire" ~ "Experiment 3",
    )
  ) |>
  summarise(
    n = n(),
    n_excluded = sum(!valid_status == "Valid"),
    across(age, lst(mean, sd, min, max)),
    l2_lst = list(l2),
    .by = c(group, exp)
  ) |>
  mutate(
    l2_lst = map(
      l2_lst,
      function(x) {
        y <- table(x)
        y <- y[names(y) != "None"]
        y <- paste0(names(y), " (", y, ")")
        return(y)
      }
    ),
    n_excluded = paste0(n, " (", n_excluded, ")"),
    age_mean_sd = paste0(round(age_mean, 2), " (", round(age_sd, 2), ")"),
    age_min_max = paste0(age_min, "--", age_max)
  ) |>
  select(group, n_excluded, age_mean_sd, age_min_max, l2_lst)

@@

In each trial, participants listened to one audio file, which contained a single word. Audio files were the same ones used in child experiments conducted in the [LABORATORY NAME REDACTED FOR ANONYMITY]. These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat [@broersma2021praat]. The average duration of the Catalan audio files was `r round(durations$cat_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$cat_eng$duration_sd, 2)`, range = `r round(durations$cat_eng$duration_min, 2)`-`r round(durations$cat_eng$duration_max, 2)`). The average duration of the Spanish audio files was `r round(durations$spa_eng$duration_mean, 2)` seconds (*SD* = `r round(durations$spa_eng$duration_sd, 2)`, range = `r round(durations$spa_eng$duration_min, 2)`-`r round(durations$spa_eng$duration_max, 2)`).

@@

```{r}
#| label: tbl-stimuli
#| tbl-cap: "Stimuli information. *M*: mean, *SD*: standard deviation, *Min*: minimum, *Max*: maximum."
tbl_data <- stimuli |>
  drop_na(freq_zipf_2, lv, neigh_n_h) |>
  summarise(
    across(c(freq_zipf_2, lv, neigh_n_h), list(mean, sd, min, max)),
    .by = c(group)
  )

@@

tbl_data |>
  tt(
    width = c(
      0.16,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07,
      0.07
    )
  ) |>
  format_tt(
    j = 2:ncol(tbl_data),
    digits = 2,
    num_zero = TRUE,
    num_fmt = "decimal"
  ) |>
  group_tt(
    j = list(
      "Lexical frequency" = 2:5,
      "Phonological similarity" = 6:9,
      "CLPN" = 10:13
    ),
  )
```

@@

$$
\begin{aligned}
&\textbf{Likelihood}  \\
y_{i} \sim & \text{Bernoulli}(p_{i}) \\
&\textbf{Parameters}  \\
\text{Logit}(p_{i}) = &  \beta_{0[p,w]} + \beta_{1[p]} \text{Frequency}_{i} + \beta_{2[p]} \text{Similarity}_i +  \beta_{3[p]} \text{CLPN}_i + \beta_{4[p]} \text{Group}_i + \\
&\beta_{5[p]} (\text{Similarity}_i \times \text{CLPN}_i) + \beta_{6[p]} (\text{Similarity}_i \times \text{Group}_i) + \\
& \beta_{7[p]} (\text{CLPN}_i \times \text{Group}_i) + \beta_{8[p]} (\text{Similarity}_i \times \text{CLPN}_i \times \text{Group}_i) \\
\beta_{0-8[p,w]} \sim & \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \text{ and  word } w \text{ in 1, ..., } W \\
\beta_{1-8[p]} \sim &  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p \text{ in 1, ..., } P \\
&\textbf{Prior}  \\
\mu_{\beta_{p,w}}  \sim &  \mathcal{N}(0, 0.1) \\
\sigma_{\beta_{p}},  \sigma_{\beta_{w}} \sim & \text{Exponential}(3) \\
\rho_{p}, \rho_{w} \sim & \text{LKJCorr}(5) \\
\end{aligned}
$$ {#eq-1}

@@

```{r}
#| label: fig-epreds-1
#| cache: false
#| fig-height: 5
#| fig-width: 10
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 1. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Predictions are plotted separately for English participants translating Catalan words, and for English participants translating Spanish words. Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
get_epreds <- function(
  model,
  data,
  n = 100,
  lv = seq(0, 1, length.out = 100),
  neigh_n_h = c(0, 2, 4, 8, 12),
  ...
) {
  lv_std <- (lv - mean(data$lv, na.rm = TRUE)) / sd(data$lv, na.rm = TRUE)
  neigh_n_h_std <- (neigh_n_h - mean(data$neigh_n_h, na.rm = TRUE)) /
    sd(data$neigh_n_h, na.rm = TRUE)
  knowledge <- unique(model$data$knowledge)
  confidence <- unique(model$data$confidence)
  freq_zipf_2_std <- 0
  group <- unique(model$data$group)
  experiment <- unique(model$data$experiment)
  nd <- expand_grid(
    freq_zipf_2_std,
    neigh_n_h_std,
    lv_std,
    knowledge,
    confidence,
    group,
    experiment
  )
  epreds <- tidybayes::add_epred_draws(nd, model, re_formula = NA, ...)
  return(epreds)
}

@@

We collected data from `r n_participants` Spanish native adults living in Spain (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants in Spain were contacted via announcements at the University campus(es), and were compensated €5 or an Amazon voucher for the same value. Participants gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the [ETHICAL COMMITTEE REDACTED FOR ANONYMITY].

@@

```{r}
#| label: fig-epreds-2
#| cache: false
#| fig-cap: "Posterior model-predicted mean accuracy in Experiment 2. Predictions were generated from 4,000 posterior samples, extracted for different values of *CLPN* (0, 2, 4, 8, 12) and *Similarity* (1-100). Lines indicate mean predictions, and intervals indicate 95%, 89%, 78%, 67%, and 50% credible intervals (CrI)."
#| fig-height: 3.5
#| fig-width: 10
lv_std <- (lv - mean(dataset_2$lv_std, na.rm = TRUE)) /
  sd(dataset_2$lv_std, na.rm = TRUE)
neigh_n_h <- c(0, 2, 4, 8, 12)
epreds_2 <- get_epreds(exp_2_m0, dataset_2, neigh_n_h = neigh_n_h, lv = lv)
epreds_2$neigh_n_h_std <- factor(
  epreds_2$neigh_n_h_std,
  levels = unique(epreds_2$neigh_n_h_std),
  labels = paste0(neigh_n_h, " neighbours"),
  ordered = TRUE
)

@@

We collected data from `r n_participants` British English native adults living in United Kingdom (*Mean* = `r round(participants_age$age_mean, 2)` years, *SD* = `r round(participants_age$age_sd, 2)`, range = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`, `r n_female[1]` female). Data collection took place from `r collection_dates[1]` to `r collection_dates[2]`. Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was approved by the [ETHICAL COMMITTEE REDACTED FOR ANONYMITY]. Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. Stimuli were the same lists of Catalan and Spanish stimuli as in Experiment 1.

@@

Phonological transcriptions like X-SAMPA are abstract representations that ignore non-phonological contrasts (e.g., phones), and consider different symbols as completely different phonemes, disregarding the fact that some phonemes share more phonetic features than others. Additionally, prosodic cues such as lexical stress, which our measure of phonological similarity does not include, might also provide participants further information about the correct translation of the presented word-forms. Altogether, it is likely that participants in the present study were able to exploit additional cues in the acoustic signal to provide correct translations. Future studies may consider a measure of phonological similarity that captures finer-grained phonetic or prosodic cues present in the acoustic signal presented to participants. explore the relative contribution of such cues in the occurrence of homophonic translation. Another possible limitation of the present study is the limited size and diversity of the all of which had high lexical frequency and low age-of-acquisition. To better characterise the factors that guide listeners’ ability to match words from an unfamiliar language with words in their native language lexicon, words with more varying levels of difficulty should be included in future studies.
